Fluent statistical computing interfaces for biological data analysis

### 2020-07-19 Rmd/05-embeddings.Rmd
update build
white space
update embeddings chapter

@@

<!--
- review of embedding algorithms, interactive graphics, tours
- technology
- design
- examples/benchmarks
  - Kevin moon
  - single cell
https://jlmelville.github.io/smallvis/init.html#other_initialization_options
-->

There has been a rapid uptake in the use of non-linear dimensionality reduction (NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in the natural sciences as part of cluster orientation and dimension reduction workflows. The appropriate use of these methods is made difficult by their complex parameterisations and the multitude of decisions required to balance the preservation of
local and global structure in the resulting visualisation. We present a visual analytics framework for the pragmatic usage of NLDR methods by combining them with a technique called the tour. A tour is a sequence of interpolated linear projections of multivariate data onto a lower dimensional space. The sequence is displayed as a dynamic visualisation, allowing a user to see the shadows the high-dimensional data casts in a lower dimensional view. By linking the tour to a view
obtained from an NLDR method, we can preserve global structure and through user interactions like spatial linked brushing observe where
the NLDR view may be misleading. We display several case 
studies from single cell genomics, that shows our approach is useful for
cluster orientation tasks. 
The implementation of our framework is available as an R package called `liminal` available at https://github.com/sa-lee/liminal.

@@

Choosing an appropriate embedding presents a challenge to the analyst. How does
an analyst know whether the embedding has captured the underlying topology and
geometry of the high dimensional space? The answer depends on the analyst's
workflow.  @Brehmer2014-hk characterised two main workflow steps that an
analyst performs when using embedding techniques: dimension reduction and
cluster orientation. The first relates to dimension reduction achieved by using
an embedding method, here an analyst wants to characterise and map meaning onto
the embedded form, for example identifying batch effects from a high throughput
sequencing experiment, or identifying a gradient or trajectory along the
embedded form @Nguyen2019-yh. The second relates to using embeddings as part of
a clustering workflow. Here analysts are interested in identifying and naming
clusters and verifying them by either applying known labels or colouring by
variables that are a-priori known to distinguish clusters. Both of these
workflow steps rely on the embedding being 'faithful' or the original high
dimensional dataset, and become much more difficult when there is no underlying
ground truth.

Embedding methods can be classified into two broad groups: linear and
non-linear methods. Linear methods perform a linear transformation of the data;
one example is principal components analysis (PCA) which performs an
eigendecomposition of the estimated sample covariance matrix. The eigenvalues
are sorted in decreasing order and represent the variance explained by each
component (eigenvector). A common approach to deciding on the number of
principal components to retain is to plot the proportion of variance explained
by each component and choose a cut-off.

Non-linear methods generally perform pre-processing on the high-dimensional data
such as generating a neighborhood graph and perform transformations on the
pre-processed form. We restrict our attention to three methods that are commonly
used in high-throughput biology: t-distributed stochastic neighbor embedding
(t-SNE), uniform manifold alignment and projection (UMAP), and potential of
heat-diffusion for affinity-based transition embedding (PHATE). The t-SNE
algorithm estimates the similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. The resulting configuration is
the one that minimizes the Kullback-Leibler divergence between the two
distributions. A recent theoretical contribution by @Linderman2019-dq proved
that t-SNE can recover spherical and well separated cluster shapes, and
proposed new approaches for tuning the optimisation parameters. It is a known
problem that t-SNE can have trouble recovering global structure and that
configurations can be highly dependent on how the algorithm is initialised and
parameterized [@Wattenberg2016-ji; @Kobak2019-lm]. UMAP is a method that is
related to LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest
neighbor graph. Its main differences are that it uses a different cost function
(cross entropy) which is optimized using stochastic gradient descent and
defines a different kernel for similarities in the low dimensional space. Due
to it's computational speed it's possible to generate UMAP embeddings in more
than three dimensions. Finally, PHATE, inspired by diffusion maps, is based on
estimating an affinity matrix via a distance matrix and k-nearest neighbors
graph. The algorithm de-noises estimated distances in high dimensional space via
transforming the affinity matrix into a Markov transition probability matrix and
diffusing this matrix over a fixed number of time steps. Then the diffused
probabilities are transformed once more to construct a distance matrix, and
multidimensional scaling is performed to construct a 2d embedding for
visualization.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly sufficient for detecting class separation, however
they noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2d scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

A complementary approach for visualizing structure in high dimensional data is
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional orthonormal basis matrix, that is represented as a
dynamic visualization. The sequence of generated bases are interpolated to form
the tour path, allowing a user to explore the subspace of projections. A grand
tour corresponds to choosing new bases at random, and can give an overview of
the structure in the data. Instead of picking projections at random, a guided
tour can be used to generate a sequence 'interesting' projections as quantified
by an index function. Given the dynamic nature of the tour, user interaction is
important for controlling and exploring the visualisation: the tour has been
used previously by @Wickham2015-cx as tool for exploring statistical model fits
and by @Buja1996-fk for exploring factorial experimental designs.

While there has been much work on the algorithmic details of the aforementioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparisons between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to their distance
in the original high-dimensional space [@Ovchinnikova2019-gf]. The work by
@Pezzotti2017-cz provides a user guided and modified form of the t-SNE
algorithm, that allows users to modified optimisation parameters in real-time.
Similarly, the embedding projector is a web interface to running UMAP, t-SNE or
PCA live in the browser and provides interactions to color points, and
highlights nearest neighbors [@Smilkov2016-hp].

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effect on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. We propose a more pragmatic workflow inspired by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks. This workflow is incorporated into an R package
called `liminal` (Available: https://github.com/sa-lee/liminal).

## Design

We propose using tours as part of an analyst's workflow in in performing
dimensionality reduction tasks. We have made extensive use of ensemble
graphics, that is aligning related plots alongside each other to provide
context. As we will see in the case studies, this allows analysts to quickly
compare views from embedding methods and allows them to see how the embedding
method alters the global structure of their data. Using ensembles allows the
use of interaction techniques, that allow analysts to perform cluster
orientation tasks via linking multiple views. This approach allows our
interface, to achieve the three principles for interactive high-dimensional
data visualisation outlined by @Buja1996-fk: finding gestalt, posing queries,
and making comparisons.

@@

We have combined the tour view in a side by side layout with 
a scatter plot view that represents the output of an embedding algorithm. 
These views are linked; analysts can brush regions or
highlight collections of points in either view. Linked highlighting can
be performed when points have been previously labeled according to some
discrete structure, i.e. cluster labels are available. This is achieved
via the analyst clicking on groups in the legend, which causes unselected groupings to have their points become less opaque. Consequently, 
simple linked highlighting can alleviate a known downfall of methods such as 
UMAP or t-SNE: that is distances between clusters are misleading. By
highlighting corresponding clusters in the tour view, the analyst can
see the relationship between clusters, and therefore obtain a more accurate 
representation of the topology of their data.

@@

As mentioned previously, when using any DR method, we are assuming
the embedding is representative of the high-dimensional dataset it was
computed from. Defining what it means for embedding to be 'representative` or
'faithful' to high-dimensional data is ill-posed and depends on the underlying
task an analyst is trying to achieve. At the very minimum, we are interested
in distortions and diffusions of the high-dimensional data. Distortions
occur when points that are near each other in the embedding view are 
far from each other in the original dataset. This implies that the embedding 
is not continuous. Diffusions occur when points are far from each other in the embedding view are near in the original data.  Whether, points are near or far 
is reliant on the distance metric used; distortions and diffusions can
be thought of as the preservation of distances or the nearest neighbors
graphs between the high-dimensional space and the embedding space. As distances
can be noisy in high-dimensions, ranks can be used instead as has been 
proposed by ... Identifying distortions and diffusions allows an analyst
to investigate the quality of their embedding and revise them iteratively. We
propose this can be done visually using our side-by-side tour and embedding
views. Some quality checks, such as finding distortions can be identified using simple linked brushing, however we can also interrogate them via spatial
brushes and brush composition.

 Look up nearest neighbours graph from points that lie in a brushing region.

Highlight the corresponding neighbours using colour or transparency in the linked view.

- The $k$ nearest neighbours graph can be pre-computed quickly, for either $X$, $Y$ or both.
- Instead of using the neighbour indices, we could use the neighbour distances instead.
- Composition of multiple brushes could be used to show where there are matches/mismatches between nearest neighbour graphs.

Using a linked neighbourhood brush, we can 
visually investigate the nearest neighbour relationships in the 
high-dimensional space via brushing in the embedding view. The user can select
the number of nearest neighbours directly, and modify the distance metric 
used for determining the neighbours. This allows users to interrogate the stability of clusters generated in the embedding view. Multiple brushes
can be used to pose queries in either the tour view or the embedding view;
interface controls allow these brushes to combine using logical operators
such as 'and', 'or', or 'not'. The use of linked brushing goes beyond
simple color highlights, allowing analysts to get a more holistic
view of the effect of an embedding algorithm.

## Implementation

We have implemnted the above design as an open source R package 
called `liminal` [@baseR]. The package allows analysts to constuct
concatenated visualisations, via the
Vega-Lite grammar of interactive graphics (using `vegawidget` package)
and provides an interface for constructing brushes and manipulating tour paths
using the `shiny` and `tourr` packages. `liminal` also provides a stand-alone
interface to the tour, in addition to the linked scatterplot interfaces 
discussed above. Below we display the `liminal` API, and our approach to
generating tour paths and user interactions.

@@

- don't need to realise the entire sequence
- instead generate new bases according to a fixed frame rate
- allows user interactions to play/pause on the current view

### Linking views via brushes

@@

### Case Study 1: Exploring tree structured data with tours and t-SNE

(ref:fake-trees) Example from PHATE

```{r fake-trees, echo = FALSE, out.width = "50%", fig.cap="(ref:fake-trees)"}
knitr::include_graphics(here::here("img", "tree.png"))
```

#### Using t-SNE

- there's a lot of parameters to tweak
- emphasises locality, distance between clusters of points can be misleading
- size of clusters can be misleading 
- may require a few different runs to capture topology of data
- if there are clusters it will find them

### Case Study 2: Clustering PBMC 10x single cell RNA-seq data

### Case Study 3: Single cell mouse retina data

>>>>

```{r setup-videos, cache = FALSE}
vid_lookup <- tibble::tribble(
  ~key, ~url, ~path,
  "gaussian",  "https://player.vimeo.com/video/439635921", here::here("video", "liminal-gaussian.mov"),
  "hierachical", "https://player.vimeo.com/video/439635905", here::here("video", "liminal-hierachical.mov"),
  "trees-01", "https://player.vimeo.com/video/439635892", here::here("video", "liminal-trees-01.mov"),
  "trees-02", "https://player.vimeo.com/video/439635863", here::here("video", "liminal-trees-02.mov"),
  "mouse-01", "https://player.vimeo.com/video/439635812", here::here("video", "liminal-mouse-01.mov"),
  "mouse-02", "https://player.vimeo.com/video/439635843", here::here("video", "liminal-mouse-02.mov")
)

display_vid_html <- function(key, lookup) {
  video <- lookup[lookup$key == key, ]
  knitr::include_url(video[["url"]])
}

disply_vid_latex <- function(key, lookup) {
    video <- lookup[lookup$key == key, ]
    glue::glue(
    "\\begin{{center}}
   \\includemedia[
    activate=pageopen,
    width=640pt,
    height=400pt,
    keepaspectratio=true,
    addresource={video},
    flashvars={{%
      src={video}
      &scaleMode=stretch}
  ]{{}}{{StrobeMediaPlayback.swf}}
  \\end{{center}}",
  video = video[["path"]]
  )
}

```

@@

There has been a rapid uptake in the use of non-linear dimensionality reduction
(NLDR) methods such as t-distributed stochastic neighbour embedding (t-SNE) in
the natural sciences as part of cluster orientation and dimension reduction
workflows. The appropriate use of these methods is made difficult by their
complex parameterisations and the multitude of decisions required to balance
the preservation of local and global structure in the resulting visualisation.
We present a visual analytics framework for the pragmatic usage of NLDR methods
by combining them with a technique called the tour. A tour is a sequence of
interpolated linear projections of multivariate data onto a lower dimensional
space. The sequence is displayed as a dynamic visualisation, allowing a user to
see the shadows the high-dimensional data casts in a lower dimensional view. By
linking the tour to a view obtained from an NLDR method, we can preserve global
structure and through user interactions like spatial linked brushing observe
where the NLDR view may be misleading. We display several case studies from
single cell transcriptomics and image processing, that shows our approach is
useful for cluster orientation tasks.  The implementation of our framework is
available as an R package called **liminal** available at
https://github.com/sa-lee/liminal.

@@

Choosing an appropriate embedding presents a challenge to the analyst. How does
an analyst know whether the embedding has captured the underlying topology and
geometry of the high dimensional space? The answer depends on the analyst's
workflow.  @Brehmer2014-hk characterised two main workflow steps that an
analyst performs when using embedding techniques: dimension reduction and
cluster orientation. The first relates to dimension reduction achieved by using
an embedding method, here an analyst wants to characterise and map meaning onto
the embedded form, for example identifying batch effects from a high throughput
sequencing experiment, or identifying a gradient or trajectory along the
embedded form [@Nguyen2019-yh]. The second relates to using embeddings as part of
a clustering workflow. Here analysts are interested in identifying and naming
clusters and verifying them by either applying known labels or colouring by
variables that are a-priori known to distinguish clusters. Both of these
workflow steps rely on the embedding being representative the original high
dimensional dataset, and becomes much more difficult when there is no underlying
ground truth.

As part of a visualization workflow, it's important to consider the perception
and interpretation of embedding methods as well. @Sedlmair2013-pn showed that
2D scatter plots were mostly sufficient for detecting class separation, however
they also noted that often multiple embeddings were required. For the task of
cluster identification, @Lewis2012-ai showed experimentally that novice users
of non-linear embedding techniques were more likely to consider clusters of
points on a 2d scatter plot to be the result of a spurious embedding compared
to advanced users who were aware of the inner workings of the embedding
algorithm.

A complementary approach for visualizing structure in high dimensional data is 
the tour. A tour is a sequence of projections of a high dimensional dataset
onto a low-dimensional basis matrix, and is represented as a
animated visualization.Given the dynamic nature of the tour, user interaction is 
important for controlling and exploring the visualisation: the tour has been
used previously by @Wickham2015-cx and @Li2020-rh as a tool for 
exploring statistical model fits and by @Buja1996-fk for exploring the space of 
factorial experimental designs.

While there has been much work on the algorithmic details of the aforementioned
embedding methods, there has been relatively few tools designed to assist users
to interact with these techniques and assist them in making comparisons between
embeddings and performing the aforementioned cluster orientation tasks. Several
interactive interfaces have been proposed for evaluating or using embedding
techniques: the Sleepwalk interface provides a click and highlight
visualisation for colouring points in an embedding according to an estimated
pair-wise distance in the original high-dimensional space [@Ovchinnikova2020-sy].
The idea behind this is that non-linear embedding methods should preserve 
locality and that points that are 'close' to each other on the embedded view
should be close according to the estimated pair-wise distance. 
Similarly, the TensorFlow embedding projector is a web interface to 
running some non-liner embedding methods live in the browser and 
provides interactions  to colour points, and select nearest neighbours
[@Smilkov2016-hp]. Finally, the work by @Pezzotti2017-cz provides a user 
guided and modified form of the  t-SNE algorithm, that allows users to modified 
optimisation parameters in real-time.

There is no one-size fits all: finding an appropriate embedding for a given
dataset is a difficult and a somewhat poorly defined problem. For non-linear
methods, there are a lot of parameters to explore that can have an effect on
the resulting visualisation and interpretation. Interfaces for evaluating
embeddings require interaction but should also be able to be incorporated into
an analysts workflow. We propose a more pragmatic workflow inspired by
incorporating interactive graphics and tours with embeddings that allows users
to see a global overview of their high dimensional data and assists them with
cluster orientation tasks. This workflow is incorporated into an R package
called **liminal** (Available: https://github.com/sa-lee/liminal).

## Dimensionality Reduction Background

In the following section, we provide a brief overview of the goals of DR
methods with respect to the data analyst, in addition to their high level the
mathematical details. Here we restrict our attention to two recent methods that
are commonly used in the literature: t-distributed stochastic neighbour embedding 
(t-SNE) and uniform manifold alignment and projection (UMAP); however we do 
mention several other techniques.

To begin we suppose the data is in the form rectangular
matrix of real numbers, $X = [\mathbf{x_1}, \dots, \mathbf{x_n}]$, where $n$ is
the number of observations in $p$ dimensions.  The purpose of any DR algorithm
is to find a low-dimensional representation of the data, 
$Y = [\mathbf{y_1}, \dots, \mathbf{y_n}]$, such that $Y$ is an $n \times d$
matrix where $d \ll p$. The hope of the analyst is that the DR procedure to 
produces $Y$ will remove noise in the original dataset while retaining any 
latent structure.

DR methods can be classified into two broad classes: linear and non-linear
methods. Linear methods perform a linear transformation of the data, that is,
$Y$ is a linear transformation of $X$; one example is principal components
analysis (PCA) which performs an eigendecomposition of the estimated sample
covariance matrix. The eigenvalues are sorted in decreasing order and represent
the variance explained by each component (eigenvector). 
A common approach to deciding on the number of principal components to retain is
to plot the proportion of variance explained by each component and choose a 
cut-off.

For non-linear methods $Y$ is generated via a pre-processed form of the input
$X$ such as the $k$-nearest neighbours graph or via a kernel transformation.
Multidimensional scaling (MDS) is a class of DR methods that aims to construct
an embedding  $Y$ such that the pair-wise distances (inner products) in $Y$ 
approximate the pair-wise distances (inner products) in $X$ 
[@Torgerson1952-am; @Kruskal1964-cz]. There are many 
variants of MDS, such as non-metric scaling which amounts to replacing
distances with ranks instead [@Kruskal1964-cw]. 
A related technique is Isomap which uses a $k$-nearest neighbour graph
to estimate the pair-wise geodesic distance of points in $X$ then uses classical 
MDS to construct $Y$ [@Silva2003-xw]. Other approaches are based on diffusion
processes such as diffusion maps [@Coifman2005-ak]. A recent example of this 
approach is the PHATE algorithm [@Moon2019-ce]. Here an affinity matrix is 
estimated via the pair-wise distance matrix and k-nearest neighbours graph of $X$. 
The algorithm de-noises estimated distances in high-dimensional space via 
transforming the affinity matrix into a Markov transition probability matrix 
and diffusing this matrix over a fixed number of time steps. Then the diffused 
probabilities are  transformed once more to construct a distance matrix, 
and classical MDS is  used to generate $Y$. A general difficulty with using 
non-linear DR methods for exploratory data analysis is selecting and tuning 
appropriate parameters. To see the extent of these choices 
we now examine the underpinnings of t-SNE and UMAP.

The t-SNE algorithm estimates
the pair-wise similarity of (Euclidean) distances of points in a high
dimensional space using a Gaussian distribution and then estimates a
configuration in the low dimensional embedding space by modelling similarities
using a t-distribution with 1 degree of freedom. There are several subtleties
to the to use of the algorithm that are revealed by stepping through its
machinery.

To begin, t-SNE transforms pair-wise distances between $\mathbf{x_i}$ and
$\mathbf{x_j}$ to similarities using a Gaussian kernel:

$$ p_{i|j} = \frac{\exp(-\lVert \mathbf{x_i - x_j} \rVert ^ 2 /
2\sigma_i^2)}{\sum_{k \ne i}\exp(-\lVert \mathbf{x_j - x_k} \rVert ^ 2 /
2\sigma_i^2)} $$

The conditional probabilities are then normalised and symmetrised to form a
joint probability distribution via averaging:

$$ p_{ij} = \frac{p_{i|j} + p_{j|i}}{2n} $$

The variance parameter of the Gaussian kernel is controlled by the analyst
using a fixed value of perplexity for all observations:

$$ \text{perplexity}_i = \exp(-\log(2) \sum_{i \ne j}p_{j|i}\log_2(p_{j|i})) $$
As the perplexity increases, $\sigma^2_{i}$ increases, until its bounded above
by the number of observations , $n-1$, in the data, corresponding to
$\sigma^2_{i} \rightarrow \infty$. This essentially turns $t-SNE$ into a
nearest neighbours algorithm, $p_{i|j}$ will be close to zero for all
observations that are not in the $\mathcal{O}(\text{perplexity}_i)$
neighbourhood graph of the $i$th observation [@Van_Der_Maaten2014-zn].

Next, the target low-dimensional space, $Y$, pair-wise distances between
$\mathbf{y_i}$ and $\mathbf{y_j}$ are modelled as a symmetric probability
distribution using a Cauchy kernel:

$$ q_{ij} = \frac{w_{ij}}{Z} \\ \text{where } w_{ij} = \frac{1}{ 1 + \lVert
\mathbf{y_i - y_j} \rVert ^ 2} \text{ and } Z = \sum_{k \ne l} w_{kl}. $$

The resulting embedding $Y$ is the one that minimizes the Kullback-Leibler
divergence between the probability distributions formed via similarities of
observations in $X$, $\mathcal{P}$ and similarities of observations in $Y$,
$\mathcal{Q}$:

$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = \sum_{i \ne j} p_{ij} \log
\frac{p_{ij}}{q_{ij}}$$

Re-writing the loss function in terms of attractive (right) and repulsive
(left) forces we obtain:

$$ \mathcal{L(\mathcal{P}, \mathcal{Q})} = -\sum_{i \ne j} p_{ij}\log w_{ij} +
\log\sum_{i \ne j} w_{ij} $$

So when the the loss function is minimised this corresponds to large attractive
forces, that is, the pair-wise distances in $Y$ are small when there are
non-zero $p_{ij}$, i.e. $x_i$ and $x_j$ are close together. The repulsive force
should also be small when the loss function is minimised, that is, when
pair-wise distances in $Y$ are large regardless of the magnitude of the
corresponding distances in $X$.

Taken together, these details reveal the sheer number of decisions that an
analyst must make. How does one choose the perplexity? How should the
parameters that control the optimisation of the loss function (done with
stochastic gradient descent), like the number of iterations, or early
exaggeration ( a multiplier of the attractive force at the beginning of the
optimisation), or the learning rate be selected?  It is a known problem that
t-SNE can have trouble recovering topology and that configurations can be
highly dependent on how the algorithm is initialised and parameterized
[@Wattenberg2016-ji; @Kobak2019-lm; @Melville2020]. If the goal is cluster
orientation a recent theoretical contribution by @Linderman2019-dq proved that
t-SNE can recover spherical and well separated cluster shapes, and proposed new
approaches for tuning the optimisation parameters. However, the cluster sizes
and their relative orientation from a $t-SNE$ view can be misleading
perceptually, due to the algorithms emphasis on locality.

Another recent method UMAP has seen a large rise in popularity (at least in
single cell transcriptomics) [@McInnes2018-co]. It is a method that is related to
LargeVis [@Tang2016-oz], and like t-SNE acts on the k-nearest neighbour graph.
Its main differences are that it uses a different cost function (cross entropy)
which is optimized using stochastic gradient descent and defines a different
kernel for similarities in the low dimensional space. Due to it's computational
speed it's possible to generate UMAP embeddings in more than three dimensions.
It appears to suffer from the same perceptual issues as t-SNE, however it
supposedly preserves global structure better than t-SNE [@Coenen2019-to].

## Visual Design

We propose using tours as part of an analyst's workflow when using embedding
methods like t-SNE. We have made extensive use of ensemble graphics, that is
aligning related plots alongside each other to provide context [@Unwin2018-zb]. 
As we will see in the case studies, this allows analysts to quickly compare 
views from embedding methods and allows them to see how the embedding method 
alters the global structure of their data. Using ensembles allows the use of 
interaction techniques, that allow analysts to perform cluster orientation 
tasks via linking multiple views. This approach allows our interface, to achieve 
the three principles for interactive high-dimensional data visualisation outlined
by @Buja1996-fk: finding gestalt, posing queries, and making comparisons.

@@

We have combined the tour view in a side by side layout with a scatter plot
view that represents the output of an embedding algorithm.  These views are
linked; analysts can brush regions or highlight collections of points in either
view. Linked highlighting can be performed when points have been previously
labelled according to some discrete structure, i.e. cluster labels are
available. This is achieved via the analyst clicking on groups in the legend,
which causes unselected groupings to have their points become less opaque.
Consequently, simple linked highlighting can alleviate a known downfall of
methods such as UMAP or t-SNE: that is distances between clusters are
misleading. By highlighting corresponding clusters in the tour view, the
analyst can see the relationship between clusters, and therefore obtain a more
accurate representation of the topology of their data.

@@

As mentioned previously, when using any DR method, we are assuming the
embedding is representative of the high-dimensional dataset it was computed
from. Defining what it means for embedding to be 'representative` or 'faithful'
to high-dimensional data is ill-posed and depends on the underlying task an
analyst is trying to achieve. At the very minimum, we are interested in
distortions and diffusions of the high-dimensional data. Distortions occur when
points that are near each other in the embedding view are far from each other
in the original dataset. This implies that the embedding is not continuous.
Diffusions occur when points are far from each other in the embedding view are
near in the original data.  Whether, points are near or far is reliant on the
distance metric used; distortions and diffusions can be thought of as the
preservation of distances or the nearest neighbours graphs between the
high-dimensional space and the embedding space. As distances can be noisy in
high-dimensions, ranks can be used instead as has been proposed by @Lee2009-zb.
Identifying distortions and diffusions allows an analyst to investigate the
quality of their embedding and revise them iteratively.

We propose this can be
done visually using our side-by-side tour and embedding views.
In the simplest case, a local continuity check can be assessed via one to
one linked brushing from the embedding to the tour view. This can be generalised further with one to many brushing, via spatial brushing. This form
of brushing acts directly on $k$-nearest neighbours ($k$-nn) graph computed from a
reference dataset, when a user brushes over a region in the embedding,
all the points that match the graphs edges are selected on the 
corresponding tour view. In place of highlighting, one could use opacity
or binned colour scales to encode distances or ranks instead of the 
neighbouring points, or by highlighting. Brush composition, such as 'and', 'or', or 'not' brushes,  could also be used to interrogate mismatches between the $k$-nn graphs estimated from both the embedding and tour views. This is left for 
future implementations of **liminal**.

## Software Infrastructure

We have implemented the above design as an open source R package called
**liminal** [@baseR]. The package allows analysts to construct concatenated
visualisations, via the Vega-Lite grammar of interactive graphics via
the **vegawidget** package [@Satyanarayan2017-gs; @Lyttle2020-hp]. It provides an interface for constructing linked
and stand alone interfaces for manipulating tour paths via the **shiny** and **tourr** [@Chang2020-bq; @Wickham2011-st]. Below we provide some background
on our implementation of the tour and user interaction.

### Tours explore the subspace of $d$-dimensional projections

Here we provide an overview of the mathematical aspects of the tour for
readers who are unfamiliar with the technique before discussing the 
computational aspects of how **liminal** interfaces with them. The tour
is emphatically not a dimension reduction technique but rather a visualisation 
technique that is grounded in mathematical theory, that is able to ascertain 
the shape and global structure of a dataset via inspection of the subspace 
generated by the set of low-dimensional projections. In fact, in some
of the case studies below, we  first sphere our data via principal 
components to reduce dimensionality of our input.

Like, when using DR techniques, the tour assumes we have a real data matrix
$X$ consisting of $n$ observations in $p$ dimensions. First, the tour generates 
a  sequence of $p \times d$ orthonormal projection matrices (bases) 
$A_{t \in \mathbb{N}}$, where $d$ is typically $1$ or $2$. For each pair of 
orthonormal bases $A_{t}$ and $A_{t+1}$ that are generated, the geodesic
path between them is interpolated to form intermediate frames, and giving the 
sense of continuous movement from one basis to another. The tour is then the 
continuous visualisation of the  projections $Y_{t} = XA_{t}$, that is the 
projection of $X$ onto $A_{t}$ as the tour path is interpolated between 
successive bases. A *grand tour* corresponds to choosing new orthonormal
bases at random; allowing a user to ascertain structure via exploring the
subspace of $d$-dimensional projections. Instead of picking projections at 
random, a *guided tour* can be used to generate a sequence 'interesting' 
projections as quantified by an index function. While **liminal** is able to 
visualise guided tours, our focus in the case studies uses the grand tour to
see global structure in the data.

@@

The process of generating successive bases and interpolating between them to  
construct intermediary frames, means the tour is a dynamic visualisation 
technique. Generally, the user would set $d=2$ the tour is visualised as
a animated scatter plot. This process of constructing bases and intermediate 
frames and visualising the resulting projection is akin to making a "flip book" 
animation. Like with a flip book, an interface to the tour requires the ability
to interact and modify it in real time. The user interface generated in **liminal**
allows a user to play, pause, and reset the tour animation,
panning and zooming to modify the scales of the plot to provide context and
click events to highlight groups of points if a labelling variable has
been placed on the legend.

These interactions are enabled by treating the basis generation as a reactive
stream. Instead of realising the entire sequence, which limits the animation
to have a discrete number of frames, new bases and their intermediate frames
are generated dynamically via pushing the current projection to the 
visualisation interface. The interface listens to events like pressing a button
or mouse-dragging and reacts by pausing the stream. This process allows the
user to manipulate the tour in real time rather than having to fix the number
of bases ahead of time. Additionally, once the user has identified an 
interesting projection or is done with the tour, the interface will return the 
current basis for use downstream.

### Linking and highlighting views via interactions

The embedding and tour views are linked together via rectangular brushes;
when a brush is active, points will be highlighted in the adjacent view. 
Because the tour is dynamic, brush events that become active will pause the
animation, so that a user can interrogate the current view. By default,
brushing on the embedding view will produce a one-to-one linking with the tour
view (the special case of $k=1$ nearest neighbours brush). In future
implementations it will be possible to shift from one-to-many linking by increasing $k$ via the 'controls' panel of the interface.  If desired, the distance metric used to estimate nearest neighbours can by chosen by the user. 
The user may provide a reference data set for computing
nearest neighbours (for example a distance matrix, or the complete data matrix)
independently of the tour or embedding views. The estimation of neighbours
is done with the `FNN` package for datasets with less than 4096 observations
and `RcppAnnoy` for larger datasets or if a user selects a non-Euclidean
distance metric [@fnn-pkg; @rcpp-annoy-pkg]. Once the user is finished 
with the interface, the brush regions are returned for downstream analysis.

@@

The next section steps through case studies cases of our approach using simulated
and real data.  The videos presented show the visual 
appearance of the **liminal** interface, and how we can interact with the tour
via the controls previously described.
In the first case study, we look at three different examples
where t-SNE is applied to three simulations where cluster structure and geometry
of the underlying data is known.

In the second case study, we apply our approach to clustering 
the mouse retina data from @Macosko2015-ot, and use apply the tour
to the process of verifying marker genes.

<!--
Finally, in the last case study we
explore a subset of the QuickDraw dataset, we use **liminal** to identify
instances of images from different labels that appear similar.
-->

### Case Study 1: Exploring simulation data with tours and t-SNE

To begin we look at simulated datasets that reproduce known facts
about the  t-SNE algorithm. Our first data set consists of five spherical 5-$d$
Gaussian clusters embedded in 10-$d$ space, each cluster has the same 
covariance matrix. We then computed a t-SNE layout with default settings 
using the **Rtsne** package [@Rtsne],
and set up the **liminal** linked interface with grand tour on the 10-$d$
observations:

```{r gaussian, results="asis", eval = knitr::is_latex_output()}
cat(display_vid_latex("gaussian", vid_lookup))
```

```{r gaussian-h, eval = knitr::is_html_output()}
display_vid_html("gaussian", vid_lookup)
```

We see in the video that t-SNE has correctly split out each cluster and
laid them out in a star like formation. This is agrees with the tour view,
where once we start the animation, the five clusters begin to appear but
generally points are more concentrated in the projection view compared to 
the t-SNE layout. This can be seen from our brushing action on the t-SNE side.

We repeat the process of simulating Gaussian clusters in the next video, 
but this time we have a cluster that has more granular structure. This time we have two
Gaussian clusters with equal covariance embedded in $10-d$, and a third cluster
with hierarchical structure. This cluster has two 3-$d$ clusters embedded in 
10-$d$, where the second cluster is subdivided into three smaller clusters, that are each equidistant from each other and have the same covariance structure.
Again t-SNE has correctly identified the sub-clusters but their relative 
locations to each other is distorted, with the orange and blue groups being
far from each other in the tour view. We see in this case that is difficult
to see the sub-clusters in the tour view, however, once we zoom and highlight
they become more apparent. When we brush the sub-clusters in the t-SNE, their
relative placement is again exaggerated, with the tour showing that they
are indeed much closer than the impression the t-SNE view gives.

```{r hierachical, results='asis', eval = knitr::is_latex_output()}
cat(display_vid_latex("hierachical", vid_lookup))
```

```{r hierachical-h, eval = knitr::is_html_output()}
display_vid_html("hierachical", vid_lookup)
```

Next explore some simulated noisy tree structured data (figure
\@ref(fig:fake-trees)), our interest here is how t-SNE visualisations break
topology of the data, and then seeing if we can resolve this by tweaking the
default parameters with reference to the global view of the data set.

(ref:fake-trees) Example high-dimensional tree shaped data, $n = 3000$ and $p = 100$. *(a)* The true data lies 2-d tree like structure consisting of ten branches. This data is available in the **phateR** package and is simulated via diffusion-limited aggregation (a random walk along the branches of the tree) with Gaussian noise added [@Moon2019-ce]. *(b)* The first two principal components, which form the initial projection for the tour, note that the backbone of the three is obscured by this view. *(c)* The default t-SNE view breaks the global structure of the tree. *(d)* Altering t-SNE using the first two principal components as the starting coordinates for the embedding, results in clustering the tree at its branching points.

```{r fake-trees, echo = FALSE, fig.align = "center", fig.cap="(ref:fake-trees)"}
knitr::include_graphics(here::here("img", "tree-view.png"))
```

First, we apply principal components and restrict the results down to the first
twelve principal components (which makes up approximately 70% of the variance
explained in the data) to use with the grand tour.

Moreover, we run t-SNE using the default arguments on the complete data (this
keeps the first 50 PCs, sets the perplexity to equal 30 and performs random
initialisation). We then create a linked
tour with t-SNE layout with **liminal** as shown in the following video:

```{r trees-01, results='asis', eval = knitr::is_latex_output()}
cat(display_vid_latex("trees-01", vid_lookup))
```

```{r trees01-h, eval = knitr::is_html_output()}
display_vid_html("trees-01", vid_lookup)
```
